Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-07-30 doc/study/study.tex
Ran aspell, VCS acronym and text on Møil architecture.

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers backed by relational databases and introduces yet
		another one, which is called \emph{Møil} and is written in Ruby
		utilizing the \ac{Rails}\footnote{\url{http://rubyonrails.org}}
		framework. It brings the handy possibility of managing the database
		schema with migrations, a lot of beautiful, responsive \acs{CRUD} and
		some other nice features (see \emph{Møil} \textref{sec:moeil:features}).
		The code is available from
		GitHub\footnote{\url{https://github.com/nning/moeil}} licensed under
		AGPLv3 \cite{agpl}.

@@

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent lookups of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased. (The actual
		accommodation of E-Mails -- another big problem with horizontal scaling
		-- can not be considered here.)

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		\emph{Møil}. An in-depth analysis of the source code can not be
		provided here. Instead, the solutions will be compared by a composed set
		of criteria. The current state of the source code was checked out on
		29\textsuperscript{th} July, 2014 from the particular version control
		systems and used for evaluation.

@@

			There is some meta-data like website, programming language, last
			update in VCS and number of commits, which is not numbered and
			compared. The used programming language would be a neutral
			attribute but gets a criterium, because before the development of
			\emph{Møil}, a decision was made against deploying further PHP
			applications to accomplish a more maintainable and secure web
			hosting environment\footnote{\urlPHPBashing}.

@@

			\subsubsection*{Features (2)}
				There are features, without which a mail server admistration
				user interface would not work at all and yet I list them,
				because it is more about analyzing, how these are implemented.
				The \crit{basic domain, mailbox and alias management}{2.1} is
				one of them. The \crit{password changing ability for
				users}{2.2} is a feature, which is also quite basic and the
				first one legitimating letting users access the administration
				frontend. Some solutions also integrate into popular
				self-hosted web mailer software for password changing. When
				users are allowed to access the administration frontend,
				\crit{access control}{2.3} or authorization is needed to
				selectively restrict the access to resources. It could also be
				possible to delegate administrative access on domains or
				resources in general to certain users.
				\s

				More mature utilities also bring useful non-basic features as
				an \crit{audit trail}{2.4.1} for example, which can be used to
				log any changes to the database with time and transacting
				account. It is also possible, the audit trail provides an undo
				functionality. In some environments users demand \crit{vacation
				mailing}{2.4.2} to automatically answer any incoming mail with
				a pre-defined text when they are vacant. Some solutions serve
				users with the possibility to activate or deactivate it and
				edit the pre-defined text. Automatic answers can also be solved
				more generic with \crit{Sieve}{2.4.3} \cite{sieve}, which is a
				server-side mail filter interface. The protocol for users to
				edit the server-side filters is called \emph{ManageSieve}
				\cite{managesieve}. Another common demand of users of a mail
				system is a \crit{spam frontend}{2.4.4} to review mails, which
				got quarantined by spam or virus checking software. (Although in
				my personal experience this false negative classification does
				occure almost never with a setup common in the Open Source
				world consisting of \emph{amavisd-new}, \emph{spamassassin} and
				\emph{clamav}.)

			\subsubsection*{Security \& robustness (3)}
			\todo{Password hashing algorithms strenghts}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \crit{Test coverage}{3.2} is important for this
				evaluation and for security purposes, because it lowers the
				probability of bugs per lines of code by trend. As
				\crit{Containability}{3.3}, the modular independence (e.g.
				independence of the web interface from the Message Store file
				system) from a security perspective is perceived.

@@

		\subsection*{The contestants}
			For introduction of "the contestants", the most important
			attributes and specialities will be listed for each. A more
			detailed comparison of the features can be found under
			\textref{sec:evaluation:overview}.

@@

				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterium; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone, was a little sloppy.

@@

				PostVis Admin is the best acknowledgement for the non-PHP
				policy it was decided on when searching for a
				\emph{postfix.admin} replacement. The code is very bad
				structured and full of redundancies. There is no \crit{schema
				normalization}{1.1} and the schema seems \crit{compatible to
				\emph{postifx.admin}}{1.2} but is likely to depart from it in
				some details. \ac{SQL} injections are findable just by checking
				for criteria like if there is an \crit{access control}{2.3}
				feature, which is basically existing but not very effective,
				because it is not implemented securely. (Any logged in user
				(also non-superadmin users) seem to be able to delete any other
				user.) The default \crit{password hashing}{3.1} algorithm is
				\texttt{md5-crypt} and there are features, which do not work
				without shell execution privileges on the mail system servers
				(e.g. \texttt{amavisd-release}, so the solution is not really
				\crit{containable}{3.3}.

@@

				The Development website was not linked correctly; the linked
				hostname did not even exist in DNS, so the URL of the VCS
				could not be determined and the 0.3-r2 version was examined.
				The code is not well modularized (almost everything is to be
				found inside \texttt{include/libratuus.php}), but it is well
				more organized (when it comes to object orientation, systematic
				"routing" and access control for example) than \emph{PostVis
				Admin}. The possibility of \ac{SQL} injections is unclear but
				there is no query sanitization in the
				\texttt{mailClass->as\_db()} function, that makes use of
				\texttt{mysql\_query()}, which in turn supports multiple queries
				configuration dependently. There is an \crit{audit
				trail}{2.4.1} feature but it supports no undo. Also the default
				\crit{password hashing}{3.1} algorithm is \texttt{md5-crypt}.

@@

				\emph{VBox.Adm} makes use of raw \ac{SQL} queries much, but
				sanitizes them. At some places (e.g. in a database maintainance
				script), the queries depend on MySQL. In others (e.g. the
				authentication), they are abstractly noted and \crit{database
				independend}{1.5}. As with most other solutions, there are no
				tests. Solely the custom SMTP proxy for spam maintainance
				integration seems to be tested a little.

@@

				The only PHP tool, which uses an \ac{ORM}. (But migrations were
				not used, yet they are supported by that \ac{ORM}.)
				\emph{ViMbAdmin} has an \crit{audit log}{2.4.1} available but
				it does not support undo. There is no real \crit{SIEVE
				support}{2.4.3}, but an access restriction if configurable for
				mailboxes on different services like SIEVE. The default
				\crit{password hashing}{3.1} algorithm is salted \texttt{MD5}
				(\texttt{md5-crypt}?) and more secure algorithms are only
				supported with shell execution of \texttt{dovecotpw}, which is
				bad for \crit{containability}{3.3}. As with many other compared
				solutions, Twitter bootstrap is used as an \ac{UI} framework,
				which supports responsive design easily but \emph{ViMbAdmin}
				seems not to make use of it conciously (there is no menu on
				mobile phones, for example).

@@

		That fact alone would maybe suffice as an argument for the development
		for yet another tool but was not as important as some other ideas.
		\emph{Møil} began as a replacement for the database schema shipped with
		\emph{postfix.admin} in a private mail system of mine. The schema was
		intended to be reimplemented normalized \cite{dbnorm} and with a
		possibility of being updated through schema migrations. Also, the
		\ac{Rails} project was intended to provide the basis for command line
		utilities to import the existent data and to perform basic
		administration tasks on domains, mailboxes and aliases. As there has
		been only the schema in use, not the whole \emph{postfix.admin} web
		frontend, there was an extra solution to generate password hashes and
		send them to an administrative contact in place. A replacement for this
		extra tool at first was the only function of the web interface part of
		\emph{Møil}. The migration opportunity was also used to change the
		password hashing algorithm from unsalted \texttt{md5} to
		\texttt{sha512-crypt}, which is substantially more secure
		\cite{missing}. (\texttt{bcrypt} was considered but unfortunately
		rejected, because it is not supported by \texttt{glibc} included in
		Debian Linux 7. Therefore, none of the compared tools supports it
		natively.)

@@

				\item[\rm Tested with postfix and dovecot]\ \\
					There are some produtive installations of \emph{Møil} in
					connection with postfix and dovecot but it could possibly
					also be used with other mailing software which provides the
					opportunity to configure the \ac{SQL} statements used to
					request necessary meta data.

@@

				\item[\rm Search]\ \\
					A search for any resources is provided, which is optionally
					based on \ac{SQL} or elasticsearch for really huge
					databases (if installed and configured to be used).
			\end{description}

@@

@@

			The application model is put into graphs in the following image. It
			shows the model classes and their relations among each other. (The
			graph is generated, so unfortunately (for now) the actual attributes
			saving a relation are missing.)

@@

			\subsubsection*{Relocations}
				The relocation model is in fact updated but actually not
				further used, currently. Whenever a \texttt{Domain} or
				\texttt{Mailbox} is renamed, it is necessary to move the
				corresponding folders in the file system of the Mail Store.
				These moves are represented as \texttt{Relocation} resources
				saving \texttt{old\_domain} and \texttt{old\_username} values,
				which can be read by a daemon (yet to be implemented) running
				on the actual Mail Store.

				\todo{Other solutions?}

@@

			& Features
			&& Basic domain, mailbox and alias management
			&& Password changing for users
			&& Access control
			&& Others
			&&& Audit trail
			&&& Vacation
			&&& Sieve
			&&& Spam frontend

>>>>

	\section*{Abstract}
		This document compares existing administration user interface solutions
		for mail servers running \emph{postfix} and backed by relational
		databases and introduces yet another one, which is called \emph{Møil}
		and is written in Ruby utilizing the
		\ac{Rails}\footnote{\url{http://rubyonrails.org}} framework. It brings
		the handy possibility of managing the database schema with migrations,
		a lot of beautiful, responsive \acs{CRUD} and some other nice features
		(see \emph{Møil} \textref{sec:moeil:features}). The code is available
		from GitHub\footnote{\url{https://github.com/nning/moeil}} licensed
		under AGPLv3 \cite{agpl}.

@@

		In bigger setups, it offers the possibility for scaling to more than
		one node (horizontal scalability) to cope with high mail and user
		loads: Several servers handling incoming mail or users reading their
		mail act as readers on one or more database servers. The performance of
		the frequent look-up of domains, mailboxes and aliases (forwardings)
		for receiving mail or user logins can be increased. (The actual
		accommodation of E-Mails -- another big problem with horizontal scaling
		-- can not be considered here.)

@@

		Although an extensive evaluation of the existing administration user
		interfaces did not happen before implementing a custom solution, some
		of the programs were still sighted and some of the following criteria
		was still considered. This more systematical and detailed analysis
		shall gain an overview and provide inspiration for the development of
		\emph{Møil}. An in-depth analysis of the source code can not be
		provided here. Instead, the solutions will be compared by a composed
		set of criteria. The current state of the source code was checked out
		on 29\textsuperscript{th} July, 2014 from the particular \acp{VCS} and
		used for evaluation.

@@

			There is some meta-data like website, programming language, last
			update in VCS and number of commits, which is not numbered and
			compared. The used programming language would be a neutral
			attribute but gets a criterion, because before the development of
			\emph{Møil}, a decision was made against deploying further PHP
			applications to accomplish a more maintainable and secure web
			hosting environment\footnote{\urlPHPBashing}.

@@

			\subsubsection*{Features (2)}
				There are features, without which a mail server administration
				user interface would not work at all and yet I list them,
				because it is more about analyzing, how these are implemented.
				The \crit{basic domain, mailbox and alias management}{2.1} is
				one of them. The \crit{password changing ability for
				users}{2.2} is a feature, which is also quite basic and the
				first one legitimating letting users access the administration
				front-end. Some solutions also integrate into popular
				self-hosted web mailer software for password changing. When
				users are allowed to access the administration front-end,
				\crit{access control}{2.3} or authorization is needed to
				selectively restrict the access to resources. It could also be
				possible to delegate administrative access on domains or
				resources in general to certain users.
				\s

				More mature utilities also bring useful non-basic features as
				an \crit{audit trail}{2.4.1} for example, which can be used to
				log any changes to the database with time and transacting
				account. It is also possible, the audit trail provides an undo
				functionality. In some environments users demand \crit{vacation
				mailing}{2.4.2} to automatically answer any incoming mail with
				a pre-defined text when they are vacant. Some solutions serve
				users with the possibility to activate or deactivate it and
				edit the pretext. Automatic answers can also be solved more
				generic with \crit{Sieve}{2.4.3} \cite{sieve}, which is a
				server-side mail filter interface. The protocol for users to
				edit the server-side filters is called \emph{ManageSieve}
				\cite{managesieve}. Another common demand of users of a mail
				system is a \crit{spam front-end}{2.4.4} to review mails, which
				got quarantined by spam or virus checking software. (Although
				in my personal experience this false negative classification
				does occur almost never with a setup common in the Open Source
				world consisting of \emph{amavisd-new}, \emph{spamassassin} and
				\emph{clamav}.)

			\subsubsection*{Security \& robustness (3)}
			\todo{Password hashing algorithms strengths}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \crit{Test coverage}{3.2} is important for this
				evaluation and for security purposes, because it lowers the
				probability of bugs per lines of code by trend. As
				\crit{Containability}{3.3}, the modular independence (e.g.
				independence of the web interface from the Message Store file
				system) from a security perspective is perceived.

@@

		\subsection*{The contestants}
			For introduction of "the contestants", the most important
			attributes and specialties will be listed for each. A more detailed
			comparison of the features can be found under
			\textref{sec:evaluation:overview}.

@@

				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterion; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone, was a little sloppy.

@@

				PostVis Admin is the best acknowledgement for the non-PHP
				policy it was decided on when searching for a
				\emph{postfix.admin} replacement. The code is very bad
				structured and full of redundancies. There is no \crit{schema
				normalization}{1.1} and the schema seems \crit{compatible to
				\emph{postfix.admin}}{1.2} but is likely to depart from it in
				some details. \ac{SQL} injections are findable just by checking
				for criteria like if there is an \crit{access control}{2.3}
				feature, which is basically existing but not very effective,
				because it is not implemented securely. (Any logged in user
				(also non-superadmin users) seem to be able to delete any other
				user.) The default \crit{password hashing}{3.1} algorithm is
				\texttt{md5-crypt} and there are features, which do not work
				without shell execution privileges on the mail system servers
				(e.g. \texttt{amavisd-release}, so the solution is not really
				\crit{containable}{3.3}.

@@

				The Development website was not linked correctly; the linked
				hostname did not even exist in DNS, so the URL of the \ac{VCS}
				could not be determined and the 0.3-r2 version was examined.
				The code is not well modularized (almost everything is to be
				found inside \texttt{include/libratuus.php}), but it is well
				more organized (when it comes to object orientation, systematic
				"routing" and access control for example) than \emph{PostVis
				Admin}. The possibility of \ac{SQL} injections is unclear but
				there is no query sanitization in the
				\texttt{mailClass->as\_db()} function, that makes use of
				\texttt{mysql\_query()}, which in turn supports multiple
				queries configuration dependently. There is an \crit{audit
				trail}{2.4.1} feature but it supports no undo. Also the default
				\crit{password hashing}{3.1} algorithm is \texttt{md5-crypt}.

@@

				\emph{VBox.Adm} makes use of raw \ac{SQL} queries much, but
				sanitizes them. At some places (e.g. in a database maintenance
				script), the queries depend on MySQL. In others (e.g. the
				authentication), they are abstractly noted and \crit{database
				independent}{1.5}. As with most other solutions, there are no
				tests. Solely the custom SMTP proxy for spam maintenance
				integration seems to be tested a little.

@@

				The only PHP tool, which uses an \ac{ORM}. (But migrations were
				not used, yet they are supported by that \ac{ORM}.)
				\emph{ViMbAdmin} has an \crit{audit log}{2.4.1} available but
				it does not support undo. There is no real \crit{SIEVE
				support}{2.4.3}, but an access restriction if configurable for
				mailboxes on different services like SIEVE. The default
				\crit{password hashing}{3.1} algorithm is salted \texttt{MD5}
				(\texttt{md5-crypt}?) and more secure algorithms are only
				supported with shell execution of \texttt{dovecotpw}, which is
				bad for \crit{containability}{3.3}. As with many other compared
				solutions, Twitter bootstrap is used as an \ac{UI} framework,
				which supports responsive design easily but \emph{ViMbAdmin}
				seems not to make use of it consciously (there is no menu on
				mobile phones, for example).

@@

		That fact alone would maybe suffice as an argument for the development
		for yet another tool but was not as important as some other ideas.
		\emph{Møil} began as a replacement for the database schema shipped with
		\emph{postfix.admin} in a private mail system of mine. The schema was
		intended to be re-implemented normalized \cite{dbnorm} and with a
		possibility of being updated through schema migrations. Also, the
		\ac{Rails} project was intended to provide the basis for command line
		utilities to import the existent data and to perform basic
		administration tasks on domains, mailboxes and aliases. As there has
		been only the schema in use, not the whole \emph{postfix.admin} web
		front-end, there was an extra solution to generate password hashes and
		send them to an administrative contact in place. A replacement for this
		extra tool at first was the only function of the web interface part of
		\emph{Møil}. The migration opportunity was also used to change the
		password hashing algorithm from unsalted \texttt{md5} to
		\texttt{sha512-crypt}, which is substantially more secure
		\cite{missing}. (\texttt{bcrypt} was considered but unfortunately
		rejected, because it is not supported by \texttt{glibc} included in
		Debian Linux 7. Therefore, none of the compared tools supports it
		natively.)

@@

				\item[\rm Tested with postfix and dovecot]\ \\
					There are some productive installations of \emph{Møil} in
					connection with postfix and dovecot but it could possibly
					also be used with other mailing software which provides the
					opportunity to configure the \ac{SQL} statements used to
					request necessary meta data.

@@

				\item[\rm Search]\ \\
					A search for any resources is provided, which is optionally
					based on \ac{SQL} or
					\emph{Elasticsearch}\footnote{\url{http://www.elasticsearch.org}}
					for really huge databases (if installed and configured to
					be used).
			\end{description}

@@

			The code is strictly modularized according to the \ac{MVC} pattern,
			like it is encouraged by \ac{Rails}.

			The model connects to the \ac{DBMS} via TCP/IP or Unix Sockets. The
			abstractly noted queries are translated to the respective \ac{SQL}
			(or NoSQL) dialect. All communication with external services like
			postfix, dovecot or the relocation daemon (see
			\textref{sec:moeil:model:relocation}) happen solely with the
			\ac{DBMS} as an intermediary.

			The view templates are written in
			Haml\footnote{\url{http://haml.info}}, styling is done in SASS and
			CoffeeScript is used for client-side dynamic functionality. Markup,
			styling and scripts are delivered to the client after being
			processed by the \ac{Rails} asset pipeline, which converts HAML to
			HTML, SASS to CSS and CoffeeScript to JavaScript. It also
			compresses the output and packs it in as few as possible files to
			keep the requests to the server low.

			Controllers and their actions are routed to by URL path patterns
			and are responsible for updating the models' state and providing
			the according data for the views.

			The search is either solved through using the \ac{DBMS} via
			\ac{SQL} or by using \emph{Elasticsearch}, which is an indexed
			search server based on \emph{Apache Lucene}.

@@

			The application model is put into graphs in the following image. It
			shows the model classes and their relations among each other. (The
			graph is generated, so unfortunately (for now) the actual
			attributes saving a relation and the cardinalities are missing.)

@@

			\subsubsection*{Relocations}
			\label{sec:moeil:model:relocation}
				The relocation model is in fact updated but actually not
				further used, currently. Whenever a \texttt{Domain} or
				\texttt{Mailbox} is renamed, it is necessary to move the
				corresponding folders in the file system of the Mail Store.
				These moves are represented as \texttt{Relocation} resources
				saving \texttt{old\_domain} and \texttt{old\_username} values,
				which can be read by a daemon (yet to be implemented) running
				on the actual Mail Store.

@@

			& Features
			&& Basic domain, mailbox and alias management
			&& Password changing for users
			&& Access control
			&& Others
			&&& Audit trail
			&&& Vacation
			&&& Sieve
			&&& Spam front-end
